The Openscapes Flywheel: A framework for managers to facilitate and scale inclusive Open science practices

### 2022-08-30 manuscript.qmd
pre-box1

Solutions to large-scale environmental and social challenges require radical collaboration that blends technology and people. The Open science movement is answering that call by transforming how we work together. While often the focus is on researchers, software developers, and data, code, and publications, the transition to Open involves investment across agencies and organizations of all kinds. One critical and often-overlooked part of transitioning to Open is that it requires support from managers and leadership, who do not need to be Open science experts but who can imagine the return on investment this new way of working provides, and support their teams through time and resources to learn new skills, take risks, and even fail through the process. In this paper we describe our work developing the Openscapes Flywheel with NASA Earth science data centers, using the flywheel concept where transformations occur from relentlessly pushing a giant, heavy flywheel that builds momentum at every turn. We offer three practical ways managers can support their teams and grow morale and technical capacity across their organizations: (1) Engage bright spots; (2) Empower a learning culture; (3) Amplify Open science leaders. Together these lessons create kinder and more inclusive organizations that are more resilient as data volumes increase and software evolve, and dispel the hero myth of science. Engaging, empowering and amplifying efforts to support people while modernizing our workforce is the way that we will meet the world's greatest challenges, from climate change to social justice to democracy. We can already see how working this way increases inclusion, reproducibility, and resilience within and across government, academic, and non-profit organizations, and are hopeful for what we can do together in growing this movement.

@@

Solutions to large-scale environmental and social challenges require radical collaboration, but have been limited by the hero model, where science is competitive, siloed, and rewards an individual hero, often a white man. While we know that science has never been done by individuals, Science perpetuated the hero myth for hundreds of years @elkins-tanton2021. Open science is a global movement that offers transparency, efficiency, reproducibility, reuse, and community with intentional focus on inclusion - the vision of a new paradigm for science @unesco2021. The Open science movement was codified by the Budapest Open Access Initiative @budapestopenaccessinitiative2001, and after decades of grassroots community building work and technology development, we are seeing the turn toward Open science with the recent publication of the UNESCO Recommendations on Open Science @unesco2021, increasing numbers of funders coordinated by the Open Research Funder Group @openres, Open access publishing (REF) and now large government institutions like NASA's Transform to Open Science @nasa2022. While Open science is not new, the transition to Open practices in a digital, data intensive world, requires investment in a new way of working @nelson2022 (REF Royal Society, 2012).

Meanwhile, at many scientific organizations, managers and leadership are facing a need to modernize - upskill the workforce, update technology, foster a more inclusive culture, support a distributed workforce, create and share reports more efficiently, transparently, and reproducibly, and steward and analyze more data, to name a few - often without a roadmap and with flat budgets. Open science offers consistency and resilience for managers, yet we are still seeing many research teams continuing to do science as usual, because they don't have the bandwidth to transition @plomp2022. Researchers feel squeezed from all sides, responsible to do more with less, while gaps widen within teams and organizations about who "is technical", or interacts with data, code, and project management. Largely, if researchers transition to Open science it is often self-taught, done on the side, as a volunteer effort and often not rewarded in traditional academic or government systems.

Transitioning towards Open science isn't just something to feel good about and it isn't all or nothing. It doesn't happen by going to a single workshop and it isn't a box to check. It is about people as well as technology - enabling people to collaborate and interoperate with technology to achieve their science goals.

@@

Openscapes Champions, our flagship program, mentors cohorts of research teams in their transition to Open and kinder science @openscapesteam2022. We have led 16 cohorts with over 112 teams (\> 520 individuals) over 2-4 months as they identify their trailhead and define their pathway together @lowndes2019. Meeting remotely twice monthly, we provide dedicated learning time for researchers to explore collaborative approaches for project management, data analysis, and stewardship, and practice incorporating them into their existing work while also bonding and reflecting with colleagues within their team and cohort. Grounded in meeting researchers where they are @lowndes2017, Openscapes Champions helps teams tackle big challenges as smaller steps. The program combines Open science and open source software with deliberate approaches for inclusion and kindness through facilitation, coaching, teaching, and community building @mayes2019, @lowndes2020a, @lakey2020, @slater2016, @edmondson1999, so that folks are empowered by daily decisions and habits for their work and collaborations that boost morale, while also being more efficient. \

With this relatively small amount of support, we have seen Champions teams take this investment and make great strides on their pathway to Open science @beck2020, @bastille2021. And long before this investment shows up in scientific publications, it shows up in learning, team morale, and resilience: sharing artwork as open educational resources reused by thousands globally @horst2022; reusing role-modeled facilitation and teaching approaches to help teams transition to remote and hybrid work and support each other during the Pandemic @fay2022; as well as preparing government reports and saving hundreds of would-be emails and would-be file versions and email attachments (Adyan Rios, personal communication); thoughtfully having difficult conversations for succession planning so that decades of knowledge and data are not lost as folks retire (Eli Holmes, personal communication). These are a few examples over a one year period - it is amazing what people can do when they are supported.\

Throughout our work, we've heard from researchers across career stages and paths that finding their team and feeling less lonely and stuck has been one of the most important parts of Openscapes Champions - away from the hero model. And just like how Open science doesn't depend on heroes, Openscapes does not either. We have developed a framework that when repeated and reinforced using the flywheel concept @collins2011, helps individuals become networked Open science leaders who support others to break down silos, make lasting, meaningful, and transformative changes within and across their organizations and amplified far beyond. This work means slowing down to speed up, which we are seeing return dividends.

@@

NASA Earth Science Data Systems missions collect Earth data, including sea ice, physical oceanography, vegetation and many other parameters @earthsciencedatasystems that are used by researchers around the world for many purposes, including pressing questions in ecology and environmental science that are part of addressing the climate crisis. Currently, NASA is migrating this data to the cloud. NASA's science data centers (Distributed Active Archive Centers, or DAACs) are supporting researchers using Earthdata to migrate their workflows to the Cloud. These are paths that have not been created before but build from existing ones.

Through a NASA grant to Openscapes, we are supporting and strengthening the community of NASA DAAC staff that are already creating cloud learning resources - a project that blends technology and people for radical collaboration @nasaopenscapesmentors. We help this community of cross-DAAC staff - the NASA Openscapes Mentors - learn existing paths and develop new paths together by identifying common priorities, needs, and gaps to fill, while testing cloud learning resources and investing in teaching. Our project increases visibility of Mentors and research teams with cloud and strengthens the community of practice around open science at NASA. In doing so, with the Mentors we are piloting and iteratively co-developing the NASA Openscapes Framework.

The NASA Openscapes Framework scales the Openscapes Approach beyond the Champions program into a movement, by identifying and supporting Open science leaders within organizations - our goal is for Openscapes to leave NASA when the project concludes, with the mentor community sustaining itself, and this only happens if the importance of their work is visible, valued, and amplified. The Framework has three components: (1) Engage bright spots to strengthen awareness and excitement; (2) Empower a learning culture to strengthen confidence and skills (3) Amplify Open science leaders to strengthen communities and the Open movement.

@@

**Welcome bright spots.** As we designed our project, we intentionally avoided coming in from the outside to change the culture. Instead, we focused on identifying and connecting the bright spots @heath2010 - people across the organization already role-modeling the desired change. We reached out to DAAC managers with a welcoming description of our project and asked them to share the opportunity with people in their group who were working with research teams moving to the Cloud and who were interested in teaching and collaborating. We then provided a short self-nomination form so that folks had agency to opt-in, and asked them to sign up as pairs and trios so they would have a team member from their DAAC to reinforce and practice what they learned with the larger cohort. Opting-in was a critical first investment from both managers and staff; this helped build agency and enthusiasm in alignment with their jobs. These nominees became the Mentor Cohort, representing five DAACs in year one, and seven in year two.

**Create space and place**. As we all had different backgrounds, different experiences and expertise, technical skills and technologies, with DAACs were in different stages of the migration to the cloud, our early conversations focused on what was unique and distinct. Part of welcoming bright spots is deep listening when folks have different vocabularies to identify a shared agenda. It took time (and gentle but consistent facilitation) for those conversations to shift from distinct to what we had in common - we found shared goals and challenges as we identified our trailhead on the path of working together.

We provided scaffolding and facilitation for regular meetings, and managed software where Mentors could collaborate. Each DAAC had a different way of working, but Mentors were all agreeable to trying to use common places. Our shared software began with Google Drive and Slack. We took time to learn how to use all software, and practice how we wanted to use it together. We also prioritized sharing stories via talks, blogs, and Twitter @nasaopenscapesmentors, so that this articulation work (REF) of community building was visible and valued, and also didn't land on the Mentors.

@@

On the technical side, many in our group had not ever accessed NASA Earthdata in the Cloud. The primary block seemed to be who would pay for access, how much will it cost to use the Cloud, and how will researchers pay for cloud computing resources? Those questions are important, but they are not the first questions for staff supporting research teams, so we took them off the table. We partnered with 2i2c @holdgraf to host a JupyterHub that is supported by AWS credits. The 2i2c JupyterHub greatly reduced "time to Cloud" for Mentors and helped them experience it for the first time and build confidence. With management support to participate in this cross-DAAC activity, job alignment, a shared interest in collaborating and learning together, along with an established space, place, and barriers removed, the group could focus on how we wanted to co-develop and teach tutorials together.

@@

**Invest in learning and trust.** To strengthen skills with teaching pedagogy for open data science, we invested in building trust through psychological safety @edmondson1999 and growth mindset @dweck2006, and gave this way of working a name: kinder science @robinson2021. We did this through art and storytelling @lowndes2020, and role-modeled facilitation including live-google docing, screensharing, and making mistakes @mayes2019, and voiced the code of conduct along with Efficiency and Inclusion tips at each meeting. We also partnered with The Carpentries @carpentries2022 for an introductory Python and GitHub workshop and Instructor Training. We were the earliest adopters of Quarto @quartodeveloperteam2022, which we learned together to make collections of Jupyter notebooks - these collections provided the dual benefit of helping identify gaps during tutorial development, as well as being a welcoming entryway for Earth science researchers in our first teaching event - the cross-DAAC Cloud Hackathon in November 2021 @podaac2021.

**Work openly - iterating, sharing the imperfect.** The Mentors combined their common skills and shared goals as they co-developed tutorials for the 2021 Cloud Hackathon. The practice of sharing imperfect work came from the foundation of learning and trust, and enabled us to iterate faster, help each other, share a common teaching style with live-coding, and see emerging ways to advocate for supporting researchers. The Mentors put into practice what they learned immediately, amplifying and continually reinforcing each other as they supported 65 participants to have their first welcoming experience in the Cloud. One example that stands out was one Mentor stepping in to teach on the first morning when another Mentor had an emergency - having built trust, camaraderie, and familiarity with the content through regular meetings, tutorial review, feedback, and teaching dry runs.

@@

**Celebrate these people and share this work - inspire the bigger movement.** Through conversations with NASA leadership, blog posts, tweets, presentations, and anywhere that we were given or could make a platform, we intentionally lifted up the Mentors and their contributions, showcasing their work and co-presenting with them. The Mentors have also co-presented together, sharing their work and how each of their organizations has benefited. Having them share the work resonates with others in their organizations, and attracts others across the NASA Earthdata community doing similar work. Uplifting their work does not dim our own contributions. It helps amplify each other, continually brightening and connecting bright spots as we all are part of and inspire the bigger movement.

@@

We have learned many lessons through our work with the Openscapes Mentors and Champions and generalized and tested them as part of our Openscapes Flywheel (fig 2) that gains momentum as it is repeated with each cycle. This guides our own work as we iterate with teams, and our goal is that others can continue to utilize it as they help their organizations transition to Open science safely and efficiently. We have been investing in Mentors from the government, like NASA, NOAA, and local state agencies, as well as academic and non-profit groups. This Flywheel is a concrete approach to help people find their pathways as they move into the Open science landscape, focusing on inclusion and community building at every step of the way. As one Mentor, Ileana Fenwick says, it's about being "the community member who you would want to meet at the beginning of your journey" ([tweet](https://twitter.com/StefanieButland/status/1552720392556466176)).

![Openscapes Flywheel for expanding the Open science landscape. ](images/flywheel-image.png){fig-align="center"}

The Openscapes Flywheel is a starting point for managers and will continue to evolve as they empower bright spots across career stages and institutions to be leaders @moore2014. Importantly, both top-down (managers) and bottom-up/grassroots (mentors) efforts should be involved, with direct communication channels between them, to strengthen and build trails together @friesz2022. The Flywheel has six practices that you can do now to move forward:

>>>>

Solutions to large-scale environmental and social challenges require radical collaboration that blends technology and people --- and the Open science movement is answering that call by transforming how we work together. While often the focus is on data, publications, code, software developers, and researchers, the transition to Open involves investment across agencies and organizations of all kinds. One critical and often-overlooked part of transitioning to Open is that it requires support from managers and leadership, who do not need to be Open science experts but who can imagine the return on investment and support their teams through time and resources to learn new skills and technologies, take risks, and even fail through the process. In this paper we describe our work developing the Openscapes Flywheel with NASA Earth science data centers, using a concept where transformations occur from relentlessly pushing a giant, heavy flywheel that builds momentum at every turn. We offer three practical ways managers can support their teams and grow morale and technical capacity across their organizations: (1) Engage bright spots; (2) Empower a learning culture; (3) Amplify Open science leaders. Together these lessons create kinder and more inclusive organizations that are more resilient as data volumes increase and software evolve, and dispel the hero myth of science. Engaging, empowering and amplifying efforts to support people while modernizing our workforce is the way that we will meet the world's greatest challenges, from climate change to social justice to democracy. We can already see how working this way increases inclusion, reproducibility, and resilience within and across government, academic, and non-profit organizations, and are hopeful for what we can do together in growing this movement.

@@

Solutions to large-scale environmental and social challenges require radical collaboration, but have been limited by the hero model, where science is competitive, siloed, and rewards an individual hero, often a white man. While we know that science has never been done by individuals, Science perpetuated the hero myth for hundreds of years @elkins-tanton2021. Open science is a global movement that offers transparency, efficiency, reproducibility, reuse, and community with intentional focus on inclusion - the vision of a new paradigm for science @unesco2021. The Open science movement was codified by the Budapest Open Access Initiative @budapestopenaccessinitiative2001, and after decades of grassroots community building work and technology development, we are seeing the turn toward Open science with the recent publication of the UNESCO Recommendations on Open Science @unesco2021, increasing numbers of funders coordinated by the Open Research Funder Group @openres, and now large government institutions supporting Open access publishing @thewhitehouse2022 and research policy, including NASA's Transform to Open Science @nasa2022. While Open science is not new, the transition to Open practices in a digital, data intensive world, requires investment in a new way of working @nelson2022 @theroyalsociety.

Meanwhile, at many scientific organizations, managers and leadership are facing a need to modernize - upskill the workforce, update technology, foster a more inclusive culture, support a distributed workforce, steward and analyze more data, and create and share reports more efficiently, transparently, and reproducibly, to name a few - often without a roadmap and with flat budgets. Open science offers consistency and resilience for managers, yet we are still seeing many research teams continuing to do science as usual, because they don't have the bandwidth to transition @plomp2022. Researchers feel squeezed from all sides, responsible to do more with less, while gaps widen within teams and organizations about who "is technical", or interacts with data, code, and project management. Largely, if researchers transition to Open science it is often self-taught, done on the side, as a volunteer effort and often not rewarded in traditional academic or government systems.

Transitioning towards Open science isn't just something to feel good about and it isn't all or nothing. It doesn't happen by going to a single workshop and it isn't a box to check. It is a practice that requires investment in people as well as technology - enabling people to collaborate and interoperate with technology to achieve their science goals.

@@

Openscapes Champions, our flagship program, mentors cohorts of research teams in their transition to Open and kinder science @openscapesteam2022. By the end of 2022 we have led 16 cohorts with over 112 teams (\> 520 individuals) over 2-4 months as they identify their trailhead and define their pathway together @lowndes2019. Meeting remotely twice monthly, we provide dedicated learning time for researchers to explore collaborative approaches for project management, data analysis, and stewardship, and practice incorporating them into their existing work while also bonding and reflecting with colleagues within their team and cohort. Grounded in meeting researchers where they are @lowndes2017, Openscapes Champions helps teams tackle big challenges as smaller steps. The program combines Open science and open source software with deliberate approaches for inclusion and kindness through facilitation, coaching, teaching, and community building @mayes2019, @lowndes2020a, @lakey2020, @slater2016, @edmondson1999, so that folks are empowered by daily decisions and habits for their work that boost morale while also being more efficient. \

With this relatively small amount of support, we have seen Champions teams take this investment and make great strides on their pathway to Open science @beck2020, @bastille2021. And long before this investment shows up in scientific publications, it shows up in learning, teamwork, and resilience: through sharing artwork as open educational resources reused by thousands globally @horst2022; reusing role-modeled facilitation and teaching approaches to help teams transition to remote and hybrid work and support each other during the Pandemic @fay2022; preparing government reports and saving hundreds of would-be emails and would-be file versions and email attachments (Adyan Rios, personal communication); and thoughtfully having difficult conversations for succession planning so that decades of knowledge and data are not lost as folks retire (Eli Holmes, personal communication). These are a few examples over a one year period - it is amazing what people can do when they are supported.\

Throughout our work, we've heard from researchers across career stages and paths that finding their team and feeling less lonely and stuck has been one of the most important parts of Openscapes Champions - away from the hero model. And just like how Open science doesn't depend on heroes, Openscapes does not either. We have developed a framework that when repeated and reinforced using the flywheel concept @collins2011, helps individuals become networked Open science leaders who support others to break down silos, make lasting, meaningful, and transformative changes within and across their organizations that is amplified far beyond. This work means slowing down to speed up, which we are seeing return dividends.

@@

NASA Earth Science Data Systems missions collect Earth data, including sea ice, physical oceanography, vegetation and many other parameters @earthsciencedatasystems that are used by researchers around the world for many purposes, including pressing questions in ecology and environmental science that are part of addressing the climate crisis. Currently, NASA is migrating this data to the cloud, and NASA's science data centers (Distributed Active Archive Centers, or DAACs) are supporting researchers using Earthdata to migrate their analytical workflows to the Cloud. These are paths that have not been created before but build from existing ones.

Through a NASA grant to Openscapes, we are supporting and strengthening the community of NASA DAAC staff that are creating cloud learning resources - a project that blends technology and people for radical collaboration @nasaopenscapesmentors. We help this community of cross-DAAC staff - the NASA Openscapes Mentors - learn existing paths and develop new paths together by identifying common priorities, needs, and gaps to fill, while testing cloud learning resources and investing in teaching. Our project increases visibility of Mentors and research teams with cloud and strengthens the community of practice around open science at NASA. In doing so, with the Mentors we are piloting and iteratively co-developing the NASA Openscapes Framework.

The NASA Openscapes Framework scales the Openscapes Approach beyond the Champions program into a movement, by identifying and supporting Open science leaders within organizations - our goal is for Openscapes to leave NASA when the project concludes, with the mentor community sustaining itself, and this only happens if the importance of their work is visible and valued. The Framework has three components: (1) Engage bright spots to strengthen awareness and excitement; (2) Empower a learning culture to strengthen confidence and skills (3) Amplify Open science leaders to strengthen the Open movement.

@@

**Welcome bright spots.** As we designed our project, we intentionally avoided coming in from the outside to change the culture. Instead, we focused on identifying and connecting the bright spots @heath2010 - people across the organization already role-modeling the desired change. We reached out to DAAC managers with a welcoming description of our project and asked them to share the opportunity with people in their group who were working with research teams moving to the Cloud and interested in teaching and collaborating. We then provided a short self-nomination form so that folks had agency to opt-in, and asked them to sign up as pairs and trios so they would have a team member from their DAAC to reinforce and practice what they learned with the larger cohort. Opting-in was a critical first investment from both managers and staff; this helped build agency and enthusiasm in alignment with their jobs. These nominees became the Mentor Cohort, representing five DAACs in year one, and seven in year two.

**Create space and place**. As we all had different backgrounds, different experiences and expertise, technical skills and technologies, with DAACs in different stages of the migration to the cloud, our early conversations tended towards what was unique and distinct. Part of welcoming bright spots is listening across different vocabularies to identify a shared agenda. It took time (and gentle but consistent facilitation) for those conversations to shift from distinct to what we had in common - we found shared goals and challenges as we identified our trailhead on the path of working together.

We provided scaffolding and facilitation for regular meetings, and managed software where Mentors could collaborate. Each DAAC had a different way of working, but Mentors were all agreeable to trying to use common places. Our shared software began with Google Drive and Slack. We took time to learn how to use all software, and practice how we wanted to use it together. We also prioritized sharing stories via talks, blogs, and Twitter @nasaopenscapesmentors, so that this articulation work of community building was visible and valued, and also didn't land on the Mentors.

@@

On the technical side, many in our group had not ever accessed NASA Earthdata in the Cloud. The primary block seemed to be who would pay for access, how much will it cost to use the Cloud, and how will researchers pay for cloud computing resources? Those questions are important, but they are not the first questions for staff supporting research teams, so we took them off the table. We partnered with 2i2c @holdgraf to host a JupyterHub that is supported by AWS credits. The 2i2c JupyterHub greatly reduced "time to cloud" for Mentors and helped them experience it for the first time and build confidence. With management support to participate in this cross-DAAC activity, job alignment, a shared interest in collaborating and learning together, along with an established space, place, and barriers removed, the group could focus on how we wanted to co-develop and teach tutorials together.

@@

**Invest in learning and trust.** To strengthen skills with teaching pedagogy for open data science, we invested in building trust through psychological safety @edmondson1999 and growth mindset @dweck2006, and gave this way of working a name: kinder science @robinson2021. We did this through art and storytelling @lowndes2020, and role-modeled facilitation including live-google docing, screensharing, and making mistakes @mayes2019, and voiced the code of conduct along with efficiency tips and inclusion tips at each meeting. We also partnered with The Carpentries @carpentries2022 for an introductory Python and GitHub workshop, as well as Instructor Training. We were the earliest adopters of Quarto @quartodeveloperteam2022, which we learned together to make collections of Jupyter notebooks - these collections provided the dual benefit of helping identify gaps during tutorial development, as well as being a welcoming entryway for Earth science researchers in our first teaching event - the cross-DAAC Cloud Hackathon in November 2021 @podaac2021.

**Work openly - iterating, sharing the imperfect.** The Mentors combined their common skills and shared goals as they co-developed tutorials for the 2021 Cloud Hackathon. The practice of sharing imperfect work came from the foundation of learning and trust, and enabled us to iterate faster, help each other, share a common teaching style with live-coding, and see emerging ways to advocate for supporting researchers. The Mentors put into practice what they learned immediately, amplifying and continually reinforcing each other as they supported 65 participants to have their first welcoming experience in the Cloud. One example that stands out was one Mentor stepping in to teach on the first morning when another Mentor had an emergency - having built trust, camaraderie, and confidence with the content through regular meetings, tutorial review, feedback, and teaching dry runs.

@@

**Celebrate these people and share this work - inspire the bigger movement.** Through conversations with NASA leadership, blog posts, tweets, presentations, and anywhere that we were given or could make a platform, we intentionally lifted up the Mentors and their contributions, showcasing their work and co-presenting with them. In co-presenting shared work, the Mentors have also highlighted how each of their organizations has benefited, which resonates with others in their organizations and attracts others across the NASA Earthdata community doing similar work. Uplifting the Mentors' work does not dim our own contributions. It helps amplify each other, continually brightening and connecting bright spots as we all are part of and inspire the bigger movement.

@@

We have learned many lessons through our work with the Openscapes Mentors and Champions and generalized and tested them as part of our Openscapes Flywheel @fig-flywheel that gains momentum as it is repeated with each cycle. This guides our own work as we iterate with teams, and our goal is that others can continue to utilize it as they help their organizations transition to Open science safely and efficiently. We have been investing in Mentors from the government, like NASA, NOAA, and local state agencies, as well as academic and non-profit groups. This Flywheel is a concrete approach to help people find their pathways as they move into the Open science landscape, focusing on inclusion and community building at every step of the way. As one Mentor, Ileana Fenwick says, it's about being "the community member who you would want to meet at the beginning of your journey" @fenwick2022 .

![Openscapes Flywheel for expanding the Open science landscape. ](images/flywheel-image.png){#fig-flywheel fig-align="center"}

The Openscapes Flywheel is a starting point for managers and will continue to evolve as they empower bright spots across career stages and institutions to be leaders @moore2014. Importantly, both top-down (managers) and bottom-up/grassroots (mentors) efforts should be involved, with direct communication channels between them, to strengthen and build trails together @friesz2022. The Flywheel has six practices building from engage, empower, and amplify that you can do now to move forward:
